---
title: |
  | Suplementary Materials for
  | Targeted quantitative metabolomics with a linear mixed-effect model for  analysis of urinary nucleosides and deoxynucleosides from bladder cancer patients before and after tumor resection
author: "| Małgorzata Artymowicz\n| Wiktoria Struck-Lewicka\n| Paweł Wiczling\n| Marcin
  Markuszewski\n| Michał J. Markuszewski\n| Danuta Siluk\n"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
  bookdown::html_document2:
    fig_caption: yes
    toc: yes
    theme: flatly
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
adress: "| Department of Biopharmaceutics and Pharmacodynamics, Medical University
  of Gdańsk, Gen. J. Hallera 107, 80-416 Gdańsk, Poland\n| Department of Urology,
  Medical University of Gdańsk, Gdańsk, Poland\n"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%')
```

# Setup

**Load packages:**

```{r message=FALSE}

library(dplyr)
library(ggplot2)
require(gridExtra)
library(brms)
library(GGally)
library(knitr)
library(tidybayes)
library(tidyr)

set.seed(10271998) ## not required but assures repeatable results
```


# Load data

The raw data comprise urine concentration measurements [$\mu$M] of selected nucleosides, deoxynucleosides and creatinine. The data was collected at 7 time point (before, 24 h, about 1, 3, 6, 9, and 12 months after the  transurethral resection of bladder tumor.

```{r}
data_wide = read.csv('RawData/data.csv') 

data_wide <- data_wide %>%
  mutate(CRE = CRE*1000)%>%
  mutate(CRE2 = CRE)
head(data_wide)

```

The data was transformed to a long format. Three variables were added: i) creatinine normalized concentrations (DVCRE), (ii) logarithm of concentrations (logDV) and (iii) logarithm of creatinine normalized concentrations (lgDVCRE):

```{r}
data_long <- tidyr::gather(data_wide, MET, DV, Pse:CRE, factor_key=TRUE)

data_long <- data_long %>%
  mutate(logDV = log(DV)) %>%
  mutate(logDVCRE = log(DV/CRE2))%>%
  mutate(DVCRE = DV/CRE2) 

```

#  Exploratory graphical analisis

## Raw data grouped by metabolite

First plot displays concentrations. The second plot displays creatinine normalized concentrations.

```{r}

new_labels = c("Pse", "U", "Ino", "3mU", "2dG", "2mG", "8OH2dG", "6mA","22dmG","5MTA","MTA", "CRE")
names(new_labels) <- c( "Pse","U","Ino","X3mU","X2dG","X2mG","X8OH2dG","X6mA","X22dmG","X5MTA","MTA","CRE")

p <- ggplot(subset(data_long, TimePoint<8))+
      geom_line(aes(x = TimePoint, y = DV, group = as.factor(ID), color="gray"))+ 
      geom_point(aes(x = TimePoint, y = DV, group = as.factor(ID), color="gray"))+ 
      geom_boxplot(aes(x = TimePoint, y = DV, fill=as.factor(TimePoint)),alpha=0.5) +
      scale_color_manual(values = "gray")+
        scale_y_continuous(trans='log10') +
        scale_x_continuous(breaks=seq(0,7,1)) +
      facet_wrap(.~as.factor(MET), scales="free", labeller = as_labeller(new_labels))+
      labs(title=element_blank(), x ="Time Points", y = "Concentrations, \u03BcM", color = "ID")+
      theme(legend.position="none")

print(p)

ggsave("Manuscript/RawData.png", plot=p,  width = 15*1.4, height = 15, units = "cm", dpi=300)

p <- ggplot(subset(data_long, TimePoint<8))+
      geom_line(aes(x = TimePoint, y = DVCRE, group = as.factor(ID), color="gray"))+ 
      geom_point(aes(x = TimePoint, y = DVCRE, group = as.factor(ID), color="gray"))+ 
      geom_boxplot(aes(x = TimePoint, y = DVCRE, fill=as.factor(TimePoint)),alpha=0.5) +
      scale_color_manual(values = "gray")+
         scale_y_continuous(trans='log10') +
        scale_x_continuous(breaks=seq(0,7,1)) +
      facet_wrap(.~as.factor(MET), scales="free", labeller = as_labeller(new_labels))+
      labs(title=element_blank(), x ="Time Points", y = "Creatinine-normalized Concentrations", color = "ID")+
      theme(legend.position="none")

print(p)
```

## Correlation plots:

There is a high correlation between metabolite concentrations. It is likely due to the variation in diuresis (urine dilution). 

```{r}

data_wide_2 <- data_long %>%
  select(-c(CRE2, Sample, logDVCRE, DVCRE, DV)) %>%
  tidyr::pivot_wider(names_from = MET, values_from = logDV)

ggpairs(data_wide_2, columns = 7:(11+7),
        labeller = "label_parsed",
        upper = list(continuous = wrap("cor", size = 2)))+
theme(axis.text = element_text(size = 5))+
theme(strip.text.x = element_text(size = 5))+
theme(strip.text.y = element_text(size = 5))


rm(data_wide_2)
```

# Data analysis

The data was analyzed using [brms package](https://paul-buerkner.github.io/brms/). It fits the Bayesian generalized (non-)linear multivariate multilevel models using Stan. The default priors were used. The Time Points 3:7 were combined assuming no changes in urine concentrations for times > 2 weeks. The model assumes fixed effects of TimePoint, MET and the interaction of MET:TIMEPoint. The Sample and ID were modeled as random effects. The variance for between patient variability was different for each MET and correlated. The variance for within subjects variability was also assumed to be different for each MET.

```{r include=TRUE, eval=FALSE}

data_long$TimePoint = as.factor(data_long$TimePoint)

levels(data_long$TimePoint) <- c(1,2,3,3,3,3,3)

data_long$Sample = as.factor(data_long$Sample)

# get_prior(bf(
#     logDV ~ TimePoint + MET + MET:TimePoint + (1+MET|ID) + (1|Sample),
#     sigma ~ MET),
#     family = student,
#     data = data_long)
```

Fit the model:  
  
```{r include=TRUE, eval=FALSE}

fit <- brm(bf(
              logDV ~ TimePoint + MET + MET:TimePoint + (1+MET|ID) + (1|Sample),
              sigma ~ MET),
  family = student,
  data = data_long,
  chains = 4, 
  cores = 4)

saveRDS(fit, file = "fit.RDS")

```

Load saved data.

```{r}

fit <- readRDS(file = "fit.RDS")

```

# Posterior predictive check

The agreement between the data and simulation confirms that model predicts the data well.

```{r}

pp_check(fit, ndraws = 100)

```

# Summary of paramters

The marginal posterior distribution of each parameter. Posterior mean, standard error, 95% credible interval and Rhat statistic for each parameter of model.

```{r}
summary(fit)
```

# Conditional effects

The following graphs display predicted concentrations for a typical subject and typical diuresis.

```{r}

plot(conditional_effects(fit, effects="TimePoint:MET"))[[1]]+
      facet_wrap(.~ MET, scales="free", labeller = as_labeller(new_labels))

```

# Visualise estimated paramters

The effect of intervention is represented as a difference in an average MET concentration (for a typical subject and typical diuresis) between $TimePoint$ 2 (or 3) and  $TimePoint$ 1. The solid line corresponds to no difference between TimePoints. The broken lines shows $\pm$ 0.223 difference corresponding to (0.8-1.25) fold change.

```{r}

new_labels = c("Pse", "U", "Ino", "3mU", "2dG", "2mG", "8OH2dG", "6mA","22dmG","5MTA","MTA", "CRE")


p1<-mcmc_plot(fit, variable = "^b_TimePoint2", regex = TRUE, transformation=exp)+
 geom_vline(xintercept = c(0.8,1.25), linetype = 3)+
  geom_vline(xintercept = c(1), linetype = 1) +
  xlim(0.4,2.1)+
  xlab("Fold change from baseline")+ 
  ggtitle("24 h post surgery")+
  scale_y_discrete(labels = rev(new_labels), limits = rev)

p2<-mcmc_plot(fit, variable = "^b_TimePoint3", regex = TRUE, transformation=exp)+
 geom_vline(xintercept = c(0.8,1.25), linetype = 3)+
geom_vline(xintercept = c(1), linetype = 1) +
  xlim(0.4,2.1)+
  xlab("Fold change from baseline")+ 
  ggtitle("> 2 weeks post surgery")+
  scale_y_discrete(labels = rev(new_labels), limits = rev) 
  
p<-grid.arrange(p1,p2,ncol = 2)

ggsave("Manuscript/Effects.png", plot=p,  width = 15*1.4, height = 15, units = "cm", dpi=300)

```

There are some large effect at $TimePoint=2$ that disappear at $TimePoints3$.

# Hypothesis testing

The probability of the 0.8-fold decrease and 1.25-fold increase for each MET and TimePoint is: 

```{r}
ans1<-hypothesis(fit, c("abs(TimePoint2)>0.223",
                        "abs(TimePoint2:METU)>0.223",
                        "abs(TimePoint2:METIno)>0.223",
                        "abs(TimePoint2:METX3mU)>0.223",
                        "abs(TimePoint2:METX2dG)>0.223",
                        "abs(TimePoint2:METX2mG)>0.223",
                        "abs(TimePoint2:METX8OH2dG)>0.223",
                        "abs(TimePoint2:METX22dmG)>0.223",
                        "abs(TimePoint2:METX5MTA)>0.223",
                        "abs(TimePoint2:METMTA)>0.223",
                        "abs(TimePoint2:METCRE)>0.223"
                        ))
ans2<-hypothesis(fit, c("abs(TimePoint3)>0.223",
                        "abs(TimePoint3:METU)>0.223",
                        "abs(TimePoint3:METIno)>0.223",
                        "abs(TimePoint3:METX3mU)>0.223",
                        "abs(TimePoint3:METX2dG)>0.223",
                        "abs(TimePoint3:METX2mG)>0.223",
                        "abs(TimePoint3:METX8OH2dG)>0.223",
                        "abs(TimePoint3:METX22dmG)>0.223",
                        "abs(TimePoint3:METX5MTA)>0.223",
                        "abs(TimePoint3:METMTA)>0.223",
                        "abs(TimePoint3:METCRE)>0.223"
                        ))


ans1$hypothesis$Hypothesis=c("abs(T2:PSU)>0.223",
                        "abs(T2:U)>0.223",
                        "abs(T2:Ino)>0.223",
                        "abs(T2:X3mU)>0.223",
                        "abs(T2:X2dG)>0.223",
                        "abs(T2:X2mG)>0.223",
                        "abs(T2:X8OH2dG)>0.223",
                        "abs(T2:X22dmG)>0.223",
                        "abs(T2:X5MTA)>0.223",
                        "abs(T2:MTA)>0.223",
                        "abs(T2:CRE)>0.223")
ans2$hypothesis$Hypothesis=c("abs(T3:PSU)>0.223",
                        "abs(T3:U)>0.223",
                        "abs(T3:Ino)>0.223",
                        "abs(T3:X3mU)>0.223",
                        "abs(T3:X2dG)>0.223",
                        "abs(T3:X2mG)>0.223",
                        "abs(T3:X8OH2dG)>0.223",
                        "abs(T3:X22dmG)>0.223",
                        "abs(T3:X5MTA)>0.223",
                        "abs(T3:MTA)>0.223",
                        "abs(T3:CRE)>0.223")
ans1
ans2
#plot(ans1)

```
There is a clear evidence the concentrations of Ino has increased and MTA has decreased 24 post surgery. 

# Variation in duresis

```{r}
fit %>%
    spread_draws(r_Sample[condition,]) %>%
    median_qi(condition_mean = r_Sample, .width = c(.95, 0.66))%>%
    ggplot(aes(y = reorder(condition,condition_mean), x = condition_mean, xmin = .lower, xmax = .upper)) +
    geom_pointinterval() +
    theme(axis.text = element_text(size = 5))+
    xlab("Change in diuresis (log scale)") +
    ylab("Sample")
  
```

# Conclusions

* The lack of control group limits the possibility to answer whether there is an effect of cancer removal on urine concentration of nucleosides and deoxynucleosides. It is possible that the observed differences are related to the surgical procedure itself. 
* MTA and Ino have the largest negative and positive effect at TimePoint2 (relative to Timepoint1). The effects are however temporary and return to pretreatment levels.
  
  
# Session info
  
```{r}

sessionInfo()

```
